Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper-methods, tensor matrices

%% THE USEPACKAGES NECESSARY FOR THIS EXAMPLE
%% NOTE THAT genetics_manu_style MUST BE CALLED AFTER mychicago
\usepackage{graphicx}
\usepackage{endfloat}
\usepackage{amsfonts}
\usepackage{mychicago}
\usepackage{subfigure}
\usepackage{genetics_manu_style}
\usepackage{authblk} % improved citations
\usepackage{setspace} %double spacing
\usepackage{lineno} % line numbers
\usepackage{booktabs} % book-quality tables
\usepackage{multicol} % list corresponding authors in two separate columns
\usepackage{hyperref} % web links

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_{1}\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_{2}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\mathbf{W}_{2}$ while feature
covariates for genotypes not covered by the incomplete predictor
($\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} \mathbf{K}^{-1}_{22} \mathbf{W}_{2}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and
where $\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values of genotypes
not equipped with the incomplete predictor.

The covariance matrix of $\boldsymbol{\epsilon}$ is $\mathbf{K}_{11} -
\mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{K}_{21} =
(\mathbf{K}^{11})^{-1}$
\cite{Legarra2009}.

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} =
\mathbf{Xb} + \mathbf{W} \boldsymbol{\alpha} + \mathbf{U} \boldsymbol{\epsilon} + \mathbf{e},
\end{equation}

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  \mathbf{y}_1 \\
  \mathbf{y}_2
 \end{bmatrix}
& =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{g}_1 \\
  \mathbf{g}_2
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
 \mathbf{X}_1 \\
 \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_2
  \boldsymbol{\alpha} + \boldsymbol{\epsilon}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{\hat{W}}_{1}\boldsymbol{\alpha} + \boldsymbol{\epsilon} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}\mathbf{W} \boldsymbol{\alpha} +
\mathbf{Z}\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e}.
\end{align}

Extending this model to hybrids with parent lines from two heterotic groups
simply requires separating the feature matrices of the two predictors into a
paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D}\mathbf{W}_{D} \boldsymbol{\alpha}_{D} + 
\mathbf{Z}_{D}\mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F}\mathbf{W}_{F} \boldsymbol{\alpha}_{F} + 
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}

@@

\begin{equation} \label{eq:breeding-values}
\hat{\mathbf{g}} = 
 \begin{bmatrix}
  \hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix}
 \hat{\boldsymbol{\alpha}}
 + 
 \begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}} 
\end{equation}

>>>>

%% THE USEPACKAGES NECESSARY FOR THIS EXAMPLE
%% NOTE THAT genetics_manu_style MUST BE CALLED AFTER mychicago
\usepackage{graphicx}
\usepackage{endfloat}
\usepackage{amsfonts}
\usepackage{mychicago}
\usepackage{subfigure}
\usepackage{genetics_manu_style}
\usepackage{authblk} % improved citations
\usepackage{setspace} %double spacing
\usepackage{lineno} % line numbers
\usepackage{booktabs} % book-quality tables
\usepackage{multicol} % list corresponding authors in two separate columns
\usepackage{hyperref} % web links
\usepackage{tensor} % tensor matrices

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let the index $C$ denote matrices and vectors pertaining to the complete
predictor and let the index $I$ denote matrices and vectors pertaining to the
incomplete predictor, respectively.

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$ while feature
covariates for genotypes not covered by the incomplete predictor
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \; \tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \; \tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}} + (\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{2}^{-1}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\tensor[_{C}]{\mathbf{K}}{_{ij}}$ are partitions of
$\tensor[_{C}]{\mathbf{K}}{}$ that correspond to $\mathbf{g}_{1}$ and
$\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and where
$\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values of genotypes
not equipped with the incomplete predictor.

The covariance matrix of $\boldsymbol{\epsilon}$ is
$\tensor[_{C}]{\mathbf{K}}{_{11}} -
\tensor[_{C}]{\mathbf{K}}{_{12}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{C}]{\mathbf{K}}{_{21}} =
(\tensor[_{C}]{\mathbf{K}}{^{11}})^{-1}$
\cite{Legarra2009}

and the covariance of $\mathbf{\hat{g}}_{1}$ is
$\tensor[_{C}]{\mathbf{K}}{_{12}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}} \;
\tensor[_{I}]{\mathbf{W}}{_{2}^{\top}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{C}]{\mathbf{K}}{_{21}}\sigma^{2}_{\alpha}$
\cite{Fernando2014}.

@@

\begin{equation} \label{eq:single-step-model}
\mathbf{y} =
\mathbf{Xb} + \tensor[_{I}]{\mathbf{W}}{} \; \tensor[_{I}]{\boldsymbol{\alpha}} + \mathbf{U} \boldsymbol{\epsilon} + \mathbf{e},
\end{equation}

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  \mathbf{y}_1 \\
  \mathbf{y}_2
 \end{bmatrix}
& =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{g}_1 \\
  \mathbf{g}_2
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
 \mathbf{X}_1 \\
 \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{C}]{\mathbf{K}}{_{12}} \; \tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
  \tensor[_{I}]{\mathbf{W}}{_{2}} \;
  \tensor[_{I}]{\boldsymbol{\alpha}} + \boldsymbol{\epsilon}  \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \; \tensor[_{I}]{\boldsymbol{\alpha}} + \boldsymbol{\epsilon} \\
  \tensor[_{I}]{\mathbf{W}}{_{2}} \; \tensor[_{I}]{\boldsymbol{\alpha}} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z} \; \tensor[_{I}]{\mathbf{W}}{} \;
\tensor[_{I}]{\boldsymbol{\alpha}} +
\mathbf{Z}\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e}.
\end{align}

Extending this model to hybrids with parent lines from two heterotic groups
simply requires separating the feature matrices of the two predictors into a
paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D} \;
\tensor[_{I}]{\mathbf{W}}{_{D}} \; \tensor[_{I}]{\boldsymbol{\alpha}}{_{D}} +
\mathbf{Z}_{D} \mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F} \;
\tensor[_{I}]{\mathbf{W}}{_{F}} \;
\tensor[_{I}]{\boldsymbol{\alpha}}{_{F}} +
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}

@@

\begin{equation} \label{eq:breeding-values}
\hat{\mathbf{g}} =
 \begin{bmatrix}
  \tensor[_{I}]{\hat{\mathbf{W}}}{_{1}} \\
  \tensor[_{I}]{\mathbf{W}}{_{2}}
 \end{bmatrix}
 \tensor[_{I}]{\hat{\boldsymbol{\alpha}}}{}
 +
 \begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix}
 \hat{\boldsymbol{\epsilon}} 
\end{equation}
